How cryptic is cryptic diversity? Machine learning approaches to classifying morphological variation in the Pacific Pond Turtle (\textit{Emys marmorata})

### 2016-03-01 doc/draft_turtle.tex
response to reviewers letter started
updated most figures. added some edits relating to those figures themselves.

\usepackage{amsmath, amsthm}
\usepackage{graphicx,hyperref}
\usepackage{microtype, parskip}
\usepackage{natbib}
\usepackage{lineno}
\usepackage[font=small]{caption}
\usepackage{subcaption, multirow, morefloats}
\usepackage{subcaption, wrapfig}
\usepackage{titlesec}
\usepackage[nottoc,numbib]{tocbibind}
\usepackage{authblk, attrib, fullpage}
\usepackage{lineno}

@@

% Why did we analyze plastron shape?
%   ammenable to 2d GMM
%   easy to compare across species

@@

% additionally
% comparison with two groups

@@

The first dataset included 354 adult \textit{E. marmorata} museum specimens; a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a}. We assigned a classification to each specimen for the different binning schemes based on geographic occurrence data recorded in museum collection archives. When precise latitude and longitude information were not available we estimated them from locality information. Because \citet{Spinks2005}, \citet{Spinks2010}, and \citet{Spinks2014} did not use vouchered specimens we were not able to directly sample individuals in their studies. Therefore our specimen classifications were based solely on the geographic information, not explicit assignment using molecular data. Because the exact barriers between different biogeographic regions are unknown and unclear, we represented each hypothesis with two schemes for a total of six different schemes. The schemes differed based on where geographic boundaries were assigned. This changes the classification of certain individuals near the boundaries between groups, providing a test of the robustness of the classification schemes. Sex information was only know for a subset of the total dataset and was not included as a predictor of classification. Sex information was used to determine if observations cluster by sex or not.

@@

\subsubsection{Supervised learning}
We used three different supervised learning, or classification, approaches: linear discriminate analysis, multinomial logistic regression, and random forests. Linear discriminate analysis, also known as canonical variate analysis, is commonly used in studies of geometric morphometric data \citep{Zelditch2004,Claude2008,Mitteroecker2011}. The other two methods, however, are not. The set of possible predictors included the centroid size of each specimen along with the value of that specimen along all 22 PC axes. In all cases, the optimal subset of PCs included in the final model was chosen via maximum within-sample receiver operating characteristic values, explained below.

@@

\begin{figure}[h]
  \centering
  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/gap_res}
  \caption{Results from PAM clustering of the Riemannian shape distance for 8 different number of clusters. Vertical lines are 1 standard deviation of the mean determined from 500 resamples.}
  \label{fig:gap}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/gap_map}
  \caption{Comparison of geographic distribution of clustered observations from the 2 clustering PAM solution. Colour and shape correspond to each of the groups. There is clearly no geographic signal in the data.}
  \label{fig:gap_map}
\end{figure}

@@

\begin{figure}[h]
  \centering
  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/var_imp}
  \caption{Variable importance from the random forest models for each of the six classification schemes. Importance is measured as the mean decrease in Gini Index, which is a measure of the strength by which that variable determines CART structure. Indices that are farther to the right indicate greater variable importance.}
  \label{fig:var_imp}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/gen_res}
  \caption{Bootstrap distributions for genearlized AUC values for each of the classification schemes. Each row corresponds to a different modeling approach: multinomial logistic regression (MLR), random forest (RF), linear discriminate analysis (LDA), and LDA using best variables from random forest. Each distribution corresponds to 1000 bootstrap replicates.}
  \label{fig:gen_hist}
\end{figure}

@@

\begin{figure}[ht]
  \centering
  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/sel_val}
  \caption{Graphical representation of the AUC values from model selection for multinomial logistic regression and linear discriminate analysis, respectively. AUC model selection is based on greatest AUC value. The horizontal axis corresponds to the cummulative number of axes included in the model of interest. A highlighted point corresponds to the AUC best model for that classification scheme.}
  \label{fig:sel}
\end{figure}

\begin{table}[ht]
  \centering
  \caption{AUC values for the best model of each classification scheme for both the observed (training) data and the generalized (testing) data. Results from all three different supervised learning approaches are shown here. AUC values range between 0.5 and 1. }
  \begin{tabular}{ l | c c | c c | c c }
    \hline
    & \multicolumn{2}{| l |}{random forest} & 
    \multicolumn{2}{| l |}{multinomial logistic regression} & 
    \multicolumn{2}{| l}{linear discriminate analysis} \\
    Scheme & Observed & Generalized & Observed & Generalized & Observed & Generalized \\ 
    \hline
    \hline
    Morph 1 & 0.63 & 0.73 & 0.75 & 0.79 & 0.75 & 0.80 \\ 
    Morph 2 & 0.61 & 0.58 & 0.76 & 0.77 & 0.76 & 0.77 \\ 
    Mito 1& 0.63 & 0.62 & 0.75 & 0.63 & 0.75 & 0.63 \\ 
    Mito 2 & 0.77 & 0.67 & 0.80 & 0.64 & 0.80 & 0.63 \\ 
    Mito 3 & 0.56 & 0.64 & 0.71 & 0.74 & 0.71 & 0.73 \\ 
    Nuclear & 0.56 & 0.67 & 0.74 & 0.62 & 0.74 & 0.77 \\ 
    \hline
  \end{tabular}
  \label{tab:comp}
\end{table}

@@

\subsection{Results from other datasets}

@@

\begin{figure}[ht]
  \centering
  \begin{subfigure}[h]{0.45\linewidth}
    \includegraphics[height = 0.3\textheight, width = \textwidth, keepaspectratio = true]{figure/trach_boot}
    \caption{\textit{Tracheyms}}
    \label{fig:trach_boot}
  \end{subfigure}
  \begin{subfigure}[h]{0.45\linewidth}
    \includegraphics[height = 0.3\textheight, width = \textwidth, keepaspectratio = true]{figure/seven_boot}
    \caption{Seven}
    \label{fig:seven_boot}
  \end{subfigure}
  \caption{Distribution of 1000 bootstrap replicates of the out-of-sample AUC values from the secondary, multi-species dataset. The AUC values are from the random forest (RF), linear discriminate analysis (LDA), and multinomial logistic regression (MLR). AUC values range between 0.5 (no better than random) and 1 (perfect classification).}
  \label{fig:other_boot}
\end{figure}

@@

        \bibitem[{{Van Bocxlaer} and Hunt(2013)}]{VanBocxlaer2013}
        {Van Bocxlaer}, B. and G.~Hunt. 2013. {Morphological stasis in an ongoing
        gastropod radiation from Lake Malawi}. Proceedings of the National Academy of
        Sciences .
        \bibAnnoteFile{VanBocxlaer2013}

@@

%\begin{table}[ht]
%  \centering
%  \caption{Results of comparisons between correctly and incorrectly classified observations from the testing data set, defined P(distance - simulated \(>\) 0). For each scheme, the classifications with at least 10 observations were tested. This was done for each of the three modeling techniques included in this study: random forest (RF), multinomial logistic regression (MLR), and linear discriminate analysis (LDA).}
%  \begin{tabular}{ l | l r r r }
%    \hline
%    Scheme & Class & RF & MLR & LDA \\
%    \hline
%    \hline
%    \multirow{2}{*}{Morph 1} & CCR & 0.77 & 1.00 & 1.00 \\ 
%    & marm & 0.87 & 0.93 & 1.00 \\ 
%    \hline
%    \multirow{2}{*}{Morph 2} & CCR & 0.88 & 1.00 & 1.00 \\ 
%    & marm & 1.00 & 1.00 & 0.98 \\ 
%    \hline
%    \multirow{2}{*}{Mito 1} & CCR & 0.94 & 0.99 & 0.98 \\ 
%    & marm & 0.99 & 0.98 & 0.47 \\ 
%    \hline
%    \multirow{2}{*}{Mito 2} & marm & 0.85 & 0.97 & 0.96 \\ 
%    & pall & 0.94 & 1.00 & 1.00 \\ 
%    \hline
%    \multirow{2}{*}{Mito 3} & 1 & 0.40 & 0.92 & 0.90 \\ 
%    & 3 & 0.97 & 0.99 & 1.00 \\ 
%    \hline
%    \multirow{2}{*}{Nuclear} & marm & 1.00 & 1.00 & 1.00 \\ 
%    & pall & 0.99 & 1.00 & 1.00 \\ 
%    \hline
%  \end{tabular}
%  \label{tab:miss_tests}
%\end{table}

>>>>

\usepackage{amsmath, amsthm}
\usepackage{graphicx,hyperref}
\usepackage{microtype, parskip}
\usepackage{natbib}
\usepackage{lineno}
\usepackage[font=small]{caption}
\usepackage{subcaption, multirow, morefloats}
\usepackage{subcaption, wrapfig}
\usepackage{rotating}
\usepackage{titlesec}
\usepackage[nottoc,numbib]{tocbibind}
\usepackage{authblk, attrib, fullpage}
\usepackage{lineno}

@@

\uppercase{Why did we analyze plastron shape?}

@@

@@

The first dataset included 354 adult \textit{E. marmorata} museum specimens; a subset of those included in \citet{Angielczyk2007}, \citet{Angielczyk2011}, and \citet{Angielczyk2013a}. We assigned a classification to each specimen for the different binning schemes based on geographic occurrence data recorded in museum collection archives. When precise latitude and longitude information were not available we estimated them from locality information. Because \citet{Spinks2005}, \citet{Spinks2010}, and \citet{Spinks2014} did not use vouchered specimens we were not able to directly sample individuals in their studies. Therefore our specimen classifications were based solely on the geographic information, not explicit assignment using molecular data. Because the exact barriers between different biogeographic regions are unknown and unclear, we represented some hypothesis with two schemes for a total of six different schemes. These schemes differed based on where geographic boundaries were assigned. This changes the classification of certain individuals near the boundaries between groups, providing a test of the robustness of the classification schemes. Sex information was only know for a subset of the total dataset and was not included as a predictor of classification. Sex information was used to determine if observations cluster by sex or not. The scheme names are as follows: Mito 1 and 2 correspond to \citet{Spinks2005}, Mito 3 corresponds to \citet{Spinks2010}, Morph 1 and Morph 2 correspond to \citet{Holland1992}, and Nuclear corresponds to \citet{Spinks2014}.

@@

\subsubsection{Supervised learning}
We used three different supervised learning, or classification, approaches: linear discriminate analysis, multinomial logistic regression, and random forests. Linear discriminate analysis, also known as canonical variate analysis, is commonly used in studies of geometric morphometric data \citep{Zelditch2004,Claude2008,Mitteroecker2011}. The other two methods, however, are not. The set of possible predictors included the centroid size of each specimen along with the value of that specimen along all 22 PC axes. In all cases, the optimal subset of PCs included in the final model was chosen via maximum within-sample receiver operating characteristic values, explained below. Additionally, all of the following models assume that each of the predictors (PC axes) are statistically independent; this is a reasonable assumption because by definition principal components are statistically independent. Using PCs as predictors in regression and classification methods is called PCA regression is a common technique for for dimensionality reduction and for accounting for correlation between variables \citep{Hastie2009}. The PCs are a coordinate system where combinations of axes and ranges of those axes can describe groupings; we are not assuming that evolution happens ``along'' PC axes \citep{Uyeda2015} but use them as a coordinate system for describing observed shapes.

@@

\begin{figure}[h]
  \centering
  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/gap_res}
  \caption{Results from PAM clustering of the Riemannian shape distance for 8 different number of clusters. Vertical lines are 1 standard deviation of the mean determined from 500 resamples. 354 points are plotted, some of which overlap.}
  \label{fig:gap}
\end{figure}

\begin{figure}[h]
  \centering
  \includegraphics[height = 0.5\textheight, width = \textwidth, keepaspectratio = true]{figure/gap_map}
  \caption{Comparison of geographic distribution of clustered observations from the 2 clustering PAM solution. Colour and shape correspond to each of the groups. There is clearly no geographic signal in the data. Points are slightly transparent as 354 points are plotted but many of them overlap.}
  \label{fig:gap_map}
\end{figure}

@@

@@

@@

@@

\begin{figure}[h]
  \centering
  \includegraphics[height = 0.8\textheight, width = 0.8\textwidth, keepaspectratio = true]{figure/var_imp}
  \caption{Variable importance from the random forest models for each of the six classification schemes. Importance is measured as the mean decrease in Gini Index, which is a measure of the strength by which that variable determines CART structure. Indices that are farther to the right indicate greater variable importance.}
  \label{fig:var_imp}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/gen_res}
  \caption{Bootstrap distributions for genearlized AUC values for each of the classification schemes. Columns correspond to different modeling approachs: multinomial logistic regression (MLR), random forest (RF), linear discriminate analysis (LDA), and LDA using best variables from random forest. Rows correspond to different classification schemes. Each distribution corresponds to 1000 bootstrap replicates.}
  \label{fig:gen_hist}
\end{figure}

\begin{figure}[ht]
  \centering
  \includegraphics[height = 0.8\textheight, width = 0.8\textwidth, keepaspectratio = true]{figure/sel_val}
  \caption{Graphical representation of the AUC values from model selection for multinomial logistic regression and linear discriminate analysis, respectively. AUC model selection is based on greatest AUC value. The horizontal axis corresponds to the cummulative number of axes included in the model of interest. A highlighted point corresponds to the AUC best model for that classification scheme.}
  \label{fig:sel}
\end{figure}

\clearpage

\begin{sidewaystable}[ht]
  \centering
  \caption{AUC values for the best model of each classification scheme for both the observed (training) data and the generalized (testing) data. Results from all three different supervised learning approaches are shown here. AUC values range between 0.5 and 1. }
  \begin{tabular}{l | r r r | r r r | r r r | r r r }
    \hline
    Scheme & \multicolumn{3}{c |}{RF} & \multicolumn{3}{c |}{MLR} & \multicolumn{3}{c |}{LDA} & \multicolumn{3}{c}{LDA + RF} \\
    & Train & Test & SD(Test) & Train & Test & SD(Test) & Train & Test & SD(Test) & Train & Test & SD(Test) \\ 
    \hline
    Mito 1 & 0.67 & 0.74 & 0.05 & 0.90 & 0.67 & 0.10 & 0.87 & 0.73 & 0.05 & 0.67 & 0.73 & 0.05 \\ 
    Mito 2 & 0.77 & 0.70 & 0.05 & 0.91 & 0.58 & 0.05 & 0.89 & 0.73 & 0.05 & 0.77 & 0.74 & 0.05 \\ 
    Mito 3 & 0.74 & 0.66 & 0.05 & 0.91 & 0.54 & 0.05 & 0.88 & 0.65 & 0.05 & 0.74 & 0.66 & 0.05 \\ 
    Morph 1 & 0.81 & 0.71 & 0.06 & 0.86 & 0.73 & 0.05 & 0.85 & 0.74 & 0.05 & 0.81 & 0.79 & 0.05 \\ 
    Morph 2 & 0.68 & 0.72 & 0.05 & 0.89 & 0.67 & 0.04 & 0.86 & 0.73 & 0.05 & 0.68 & 0.72 & 0.05 \\ 
    Nuclear & 0.64 & 0.63 & 0.06 & 0.84 & 0.59 & 0.06 & 0.82 & 0.58 & 0.06 & 0.64 & 0.58 & 0.08 \\ 
   \hline
\end{tabular}
  \label{tab:comp}
\end{sidewaystable}

\clearpage
\subsection{Results from other datasets}

@@

\begin{figure}[ht]
  \centering
  \begin{subfigure}[h]{0.8\linewidth}
    \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/trach_boot}
    \caption{\textit{Tracheyms}}
    \label{fig:trach_boot}
  \end{subfigure}
  \\
  \begin{subfigure}[h]{0.8\linewidth}
    \includegraphics[height = 0.8\textheight, width = \textwidth, keepaspectratio = true]{figure/seven_boot}
    \caption{Seven}
    \label{fig:seven_boot}
  \end{subfigure}
  \caption{Distribution of 1000 bootstrap replicates of the out-of-sample AUC values from the secondary, multi-species dataset. The AUC values are from the random forest (RF), linear discriminate analysis (LDA), and multinomial logistic regression (MLR). AUC values range between 0.5 (no better than random) and 1 (perfect classification).}
  \label{fig:other_boot}
\end{figure}

@@

    \bibitem[{{Van Bocxlaer} and Hunt(2013)}]{VanBocxlaer2013}
    {Van Bocxlaer}, B. and G.~Hunt. 2013. {Morphological stasis in an ongoing
    gastropod radiation from Lake Malawi}. Proceedings of the National Academy of
    Sciences 110:13892--13897.
    \bibAnnoteFile{VanBocxlaer2013}

    \bibitem[{Uyeda et~al.(2008)Uyeda, Caetano, and Pennell}]{Uyeda2015}
    Uyeda, Josef C., D.~S.~Caetano and M.~W.~Pennell, {Comparative Analysis of 
    Principal Components Can be Misleading}. Systematic Biology 64:677--689.
    \bibAnnoteFile{Uyeda2015}

@@
